Grade Retention and Mixed Race

### 2022-04-18 products/paper/main.qmd
Re-organize discussion of example figure in methods section
I now illustrate the figure using the raw percents from Table 1 prior to discussing the models which I think is a more natural flow

To measure grade retention, I compare the age of respondents to their current grade in school. Because the ACS lacks data on the exact birthday of respondents, it is not possible to exactly identify students who are below the expected grade level. All grades include two expected age groups and a student of a given age could always reasonable be expected to be in two different grades. For example, a first grader is expected to be either six or seven years of age, while a second grader is expected to be either seven or eight years of age. Therefore, with information only on current grade and age in years, it is impossible to determine whether a seven year old in first grade has been held back or is at the expected grade level.

@@

The analysis is driven by a set of logit models that predict the likelihood of a student being clearly behind grade. For the baseline model, I consider several characteristics that may vary across groups and thus need to be controlled in all models. First, states dictates educational policy and thus can differ substantially in the likelihood of grade retention so I include fixed effects for state of residence in all analysis. I cannot identify specific school districts, but I also include dummy variables indicating whether the student lived in a central city, suburban, or rural area.[^2]

[^2]: The location cannot be determined for all students and I therefore also include a fourth category of unknown in all models.

@@

To illustrate via an example, I take the raw percent of students clearly behind grade for White, Black, and Black/White students. These percents are 3.42%, 5.30%, and 3.64%. I then display these probabilities in @fig-example. This figure clearly shows the placement of Black/White students (shown in red) relative to White and Black students. I show uncertainty for all three cases with 83.4% confidence intervals. I use 83.4% confidence intervals because overlap on such intervals indicates that the difference between the two measures is statistically distinguishable from zero at $p<.05$ (CITE). This is not true of overlap with 95% confidence intervals. The lack of overlap of the Black/White probability with either the White or Black bands indicates that it is statistically distinguishable from both monoracial probabilities.

Additionally, I include a measure of the "halfway" point between the two racial groups.[^3] This allows me to additionally determine where the biracial group falls relative to the expecation of being halfway between the two constituent monoracial groups. In this case, we can see that Black/White students are much closer to White students in their risk of grade retention, and have grade retention rates much lower than bpth Black students and the halfway expectation.

[^3]: The standard error of this halfway estimate used to construct the confidence interval is given by $\sqrt{(s_1^2+s_2^2)/2}$.

>>>>

To measure grade retention, I compare the age of respondents to their current grade in school. Because the ACS lacks data on the exact birthday of respondents, it is not possible to exactly identify students who are below the expected grade level. All grades include two expected age groups and a student of a given age could always reasonable be expected to be in two different grades. For example, a first grader is expected to be either six or seven years of age, while a second grader is expected to be either seven or eight years of age. Therefore, with information only on current grade and age in years, it is impossible to determine whether a seven year old in first grade is behind the expected grade level.

@@

@tbl-descriptive shows the percent of each racial group that is clearly behind the expected grade level. My goal is understand the level of grade retention for each biracial group relative to its constituent monoracial groups (e.g. a Black/White student compared to White and Black students). Throughout this article, I use a visual approach to illustrate this placement.

As an example, I display the percent of students clearly behind grade for White, Black, and Black/White students in @fig-example. The point estimate and confidence interval for the Black/White group is shown in red, while the two monoracial groups are shown via grey dots and confidence bands. For all three cases, I use 83.4% confidence intervals, because overlap on such intervals indicates that the observed difference between the two measures in the sample is not statistically distinguishable from zero in the population at $p<.05$ (CITE).[^2] In this case, the lack of overlap between the Black/White probability and either the White or Black band indicates that the Black/White probability of clearly being behind grade level is statistically distinguishable from both monoracial probabilities.

[^2]: While 95% confidence intervals are more common in practice, the overlap of such intervals does not necessarily indicate a lack of statistical significance

Additionally, I include a measure of the "halfway" point between the two racial groups.[^3] This halfway point allows me to determine where the biracial group falls relative to the expecation of being halfway between the two constituent monoracial groups. In this case, Black/White students are much closer to White students in their risk of grade retention, and have grade retention rates much lower than both Black students and the halfway expectation.

[^3]: The standard error of this halfway estimate used to construct the confidence interval is given by $\sqrt{(s_1^2+s_2^2)/2}$ where $s_1$ and $s_2$ are the standard errors of the estimates for the two monoracial groups.

@fig-example

The analysis is driven by a set of logit models that predict the likelihood of a student being clearly behind grade. For the baseline model, I consider several characteristics that may vary across groups and thus need to be controlled in all models. First, states dictates educational policy and thus can differ substantially in the likelihood of grade retention so I include fixed effects for state of residence in all analysis. I cannot identify specific school districts, but I also include dummy variables indicating whether the student lived in a central city, suburban, or rural area.[^4]

[^4]: The location cannot be determined for all students and I therefore also include a fourth category of unknown in all models.

@@
